\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Edits for snow

where three types of random fields are now estimated: the spatial intercept
($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and
spatially varying NAO coefficients ($\zeta_{\bm{s}}$). The NB2 distribution is
specified with a mean $\mu_{\bm{s},t}$ and size parameter $\phi$. The
observation variance scales quadratically with the mean: $\mathrm{Var}[y] = \mu
+ \mu^2 / \phi$ \citep{hilbe2011}. The $\alpha_t$ represent IID random
intercepts by year: $\alpha_t \sim \operatorname{Normal} \left(0,
\sigma_\alpha^2 \right)$.

@@

The spatially varying coefficient effect at any point in space is the
combination of the main effect and spatial random effect for \texttt{nao}. We
illustrate deriving uncertainty on their combination by simulating from the
parameter precision matrix assuming the joint parameter covariance matrix in
multivariate normal. We do this by specifying a non-null number of simulation
draws to \texttt{nsim} in \texttt{predict.sdmTMB()}. By default, \texttt{nsim >
0} will return a matrix of draws from the overall prediction. Here, we instead
   specify that we want to return draws from the \texttt{zeta\_s} random field,
   which is the spatially varying random field. This returns a matrix where each
   row matches a row of \texttt{newdata} and each column is a simulation draw.
   We then use the function \texttt{spread\_sims()} to draw 200 simulations from
   the parameters themselves, combine the main effect draws (\texttt{sims\$nao})
   with the random field values \texttt{zeta\_s}, and calculate the median for
   each row of data so we can plot them.

@@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5>>=
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
@

@@

TODO: BEFORE OR AFTER EXAMPLES

>>>>

where three types of random fields are now estimated: the spatial intercept
($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and
spatially varying NAO coefficients ($\zeta_{\bm{s}}$). The NB2 distribution is
specified with a mean $\mu_{\bm{s},t}$ and size parameter $\phi$. The
observation variance scales quadratically with the mean: $\mathrm{Var}[y] = \mu
+ \mu^2 / \phi$ \citep{hilbe2011}. The $\alpha_t$ represent IID random
intercepts by year: $\alpha_t \sim \operatorname{Normal} \left(0,
\sigma_\alpha^2 \right)$. We can then fit this model:

@@

The spatially varying coefficient effect at any point in space is the
combination of the main effect and spatially varying random effect for
\texttt{nao}. Mean estimates of the spatially varying random effect are
available in the output of \texttt{predict.sdmTMB()} in a column starting with
\texttt{zeta\_s} (here, \texttt{zeta\_s\_nao}). However, we wish to combine the
fixed and random effects with uncertainty on the combined predictions. We
illustrate a way of accomplishing this by simulating from the parameter
precision matrix assuming the joint parameter covariance matrix is multivariate
normal. We do this by specifying a non-null number of simulation draws to
\texttt{nsim} in \texttt{predict.sdmTMB()}. By default, \texttt{nsim > 0} will
return a matrix of draws from the overall prediction. Here, we instead specify
that we want to return draws from the \texttt{zeta\_s} ($\zeta_{\bm{s}}$) random
field, which is the spatially varying random field. This returns a matrix where
each row matches a row of \texttt{newdata} and each column is a simulation draw.
We then use the function \texttt{spread\_sims()} to draw 200 simulations from
the parameters themselves, combine the main effect draws (\texttt{sims\$nao})
with the random field values \texttt{zeta\_s}, and calculate the median for each
column of data so we can plot them.

@@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5.75>>=
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
@

@@
